Software Product Line Analysis Techniques

### 2018-03-05 doc/quals/quals.tex
Conclusion

@@

\section{Introduction}
A Software Product Line (SPL) is a family of related software products built from a common set of artifacts (e.g., design/specification models, source code, test cases). An SPL is composed of a set of features. A feature is an externally visible property, aspect or quality of a software system. An instance (product) of an SPL is composed of a subset of the SPL features. Not all combinations of features are valid though. A Feature Model (FM) - also known as a Variability Model (VM) - is a specification of the valid combinations of SPL features. An FM can be expressed as a boolean formula over features, or graphically as a Feature Diagram (FD).

@@

This position paper covers some research work attempting to solve this problem. One paper compares brute force (product-based), variability-aware (family-based) and sampling techniques\cite{Apel:2013}. Another extends this work to more elaborate sampling techniques and more general categories of analysis\cite{Liebig:2013}. At the theoretical end of the spectrum, one paper provides an elaborate mathematical foundation for lifting abstract interpretation based analyses to SPLs\cite{Midtgaard:2015}. At the other end, some papers provide pragmatic implementations of source code syntax checkers\cite{Gazzillo:2012}, type checkers\cite{Kastner:2012}, data flow analyses\cite{Bodden:2013}, and model checkers\cite{Classen:2013}.

@@

In \cite{Midtgaard:2015}, a systematic approach to lift abstract interpretation analyses from single programs to SPLs is presented. The paper provides a rigorous  formal treatment of abstract interpretation applied to a simple imperative language, with variability at the statement level. Formal proofs of the soundness of the lifting approach are also presented.

Given a programming language $\lang$, starting with an abstract interpretation (collecting semantics, an abstract domain, a Galois connection connecting the two, and an analysis function), the paper presents the \emph{lift} combinator, which lifts each of those components to tuples of values, with each value corresponding to a product configuration in the configuration space $\configSpace$. If $\lang$ is extended with static conditionals (e.g., CPP \#if) into a variability-aware language  $\lang'$, and the semantics are respectively extended, the \emph{lift} combinator can take a product-level abstract interpretation analysis and generate a variability-aware analysis out of it.

The formulation presented lifts single values to tuples of values, each tuple $|\configSpace|$ in length. $|\configSpace|$ is exponential in the number of features in the feature space $\featSpace$, so implementing lifted analyses this way would be intractable. Caching and memoization of intermediate values can save some computational cycles, but still the overall computational complexity of the lifted analysis will grow exponentially with the number of features. Two optimizations are proposed in the paper: \emph{value sharing} and \emph{variational abstraction}.

\emph{Value sharing} attempts to eliminate redundancy within a lifted tuple by simply mapping individual values to presence conditions, instead of enumerating the values belonging to each of the configurations. This way if all configurations share the same values except for one, the tuple would only have two values with a presence condition each. The shorter the tuple, the faster the analysis since it will need to process less intermediate values. Presence conditions can be efficiently encoded and checked for satisfiability as CNF/DNF formulas or BDDs.

\emph{Variational abstraction} pushes abstraction up to the configuration space. Instead of lifting the exact product-level analysis up to the SPL, the lifted analysis can be abstracted to an over-approximation. Pretty much like all abstraction techniques, this trades accuracy for performance. \emph{Variational abstraction} is implemented the same way as other abstraction stages, starting with an abstract domain and a Galois connection together with all the corresponding proofs. Abstraction can vary from minimal ($\alpha_\featSpace$ where no abstraction takes place) to maximal ($\alpha_\phi$ where the configuration space is collapsed into a single value). Since Galois connections compose (the composition of two Galois connections is a Galois connection), \emph{variational abstraction} fits naturally in the presented calculational abstract interpretation framework.

This lifting approach was evaluated on a single analysis (reaching definitions). Three Java SPLs (Graph PL (GPL), Prevayler and BerkeleyDB) were used. However, the evaluation numbers presented are not aggregated across the 3 codebases, but rather for only 4 source code methods (2 from GPL, 1 from Prevayler and 1 from BerkeleyDB). The configuration space of each of those methods varies between 8 and 106 configurations.

The performance (running time) of three SPL analysis techniques is measured: brute force (per product analysis), lifted analysis with sharing, and lifted analysis with maximum variational abstraction (abstracting configuration variability away). Comparing against brute force analysis, speedup due to lifting with maximum variational abstraction grows exponentially with respect with the number of configurations, and almost identical to the average analysis performance of a single configuration. The speedup due to lifting with sharing also grows exponentially, but at a slower rate. Generalizing from those performance numbers is risky though because of highly restricted sample space (a single analysis on only 4 source code methods).

Classen et al.\cite{Classen:2013} address the problem of model checking variability-aware systems. The paper introduces the Featured Transition System (FTS) formalism of models, the feature Linear Temporal Logic (fLTL) for expressing temporal logic properties, and the SNIP model checker that checks fLTL properties against FTS models.

The FTS model checking algorithm extends the state reachability algorithm for LTSs with the identification of the sets of products in which each state is reachable from an initial state. Sets ot products are defined in terms of feature expressions, which are encoded as BDDs. The FD of the model is used to exclude invalid products. Since only a single counterexample is needed whenever a product violates the property, the search can be pruned for products where a counterexample has already been found.

@@

In \cite{Kastner:2012}, a Colored Featherweight Java (CFJ) type checker is formalized and certified using the Coq\cite{Bertot:2010} proof assistant. Featherweight Java (FJ) is a simplistic subset of Java, leaving out constructs such as assignment (mutation) and interfaces. CFJ adds IDE-assisted variability to FJ, where different source code features have different colors in the IDE. This type checker is further extended to fully cover Java constructs.

@@

In \cite{Apel:2013}, product-based, sample-based and family-based model-checking strategies of SPLs are compared. Product-based (generating all products and checking each one of them individually) is used as a baseline. Sample-based strategies only generate a sample of products (single-wise, pair-wise and triple-wise in terms of valid feature combinations). Finally, family-based checking checks the whole SPL in one shot, treating variability points as conditionals.

In this work the sampling strategies used are too primitive. Systems defects that are due to the interaction of more than 3 different features (which are not rare) will not be detected by the strategies presented in this paper, simply because this work only samples up to triple-wise feature combinations.

One caveat that is not explicitly mentioned in the paper is that this family-based approach is not applicable to annotative SPLs. The case studies presented are all generative (each feature is implemented separately and then composed into an SPL). Different function implementations together with the dispatcher function are generated as a part of the SPL generation. However, annotative SPLs are less strict when it comes to variability points (they don't all map to conditionals). Late Splitting and Early Joining can't be directly implemented in CPP-based SPLs for example.

The performance of the different strategies is compared in terms of their speedup with respect to the product-based baseline. In product-based and sample-based strategies, model checking is terminated once a defect is found in a product. However, in the family-based strategy it is not possible to terminate checking when a defect is found, simply because that defect does not necessarily exist in all products. This is a threat to validity of the performance numbers, not mentioned in the paper. The family-based strategy effectively tries to finds all defects in each product in the SPL, while product-based and strategy-based strategies only find the first defect (if exists). Comparing those two different metrics is thus misleading.

\section{Scalable Analysis of Variable Software}
In \cite{Liebig:2013}, variability-aware (family-based) static analyses are compared to some sample-based techniques. Two variability-aware C language analyses are implemented and evaluated: type checking and liveness analysis. Three relatively big open source systems were used for evaluation: the Linux kernel, the Busybox codebase, and the OpenSSL library.

In addition to focusing on static analysis as opposed to model checking, there are two main differences between this work and that of \cite{Apel:2013}. First, the variability-aware analyses are applied to annotative SPLs (using the CPP to incorporate variability). Because annotative SPLs and more permissive than generative SPLs in terms of syntactic variability, some building blocks had to be purposely designed to be variability-aware. Those include variability-aware Abstract Syntax Trees (ASTs), Control Flow Graphs (CFGs) and symbol tables. In addition, a dedicated variability-aware parser (incorporating CPP logic) is used.

The second difference with respect to \cite{Apel:2013} is that the sampling techniques used in the comparison are more elaborate. In addition to pair-wise sampling (covered in \cite{Apel:2013}), single configuration, random and coverage-based sampling were also evaluated.

Similar to \cite{Apel:2013}, the main principles guiding the design of the variability-aware analyses are late splitting and early joining. These principles maximize reuse of intermediate analysis information across different variants, eventually resulting in variability-aware analyses outperforming brute-force per-product analysis by orders of magnitude. A third principle implicit in \cite{Apel:2013} but explicitly spelled out here is local variability representation. This is the mechanism of maintaining variability within intermediate analysis values.

In \cite{Bodden:2013}, inter-procedural data flow analysis (based on the IFDS \cite{Reps:1995} framework) is extended to annotative product lines. In IFDS, inter-procedural data flow analysis problems are encoded as graph reachability problems. The main idea behind lifting IFDS analyses is to guard data-flow properties of a program statements by presence conditions. This fits well within the IFDS framework because data flow properties are boolean expressions.

This work was evaluated on 3 lifted analyses: Possible Types, Reaching Definitions and Uninitialized Variables. 4 real world SPLs were used as case studies: BerkeleyDB, GPL, Lampiro and MM08. Performance results show that the lifted analyses run in minutes (in some cases seconds), while brute force analysis would take years for some of the case studies. One thing to note though is that the correctness of the lifting approach is shown empirically rather than with a logical proof.

\section{SuperC}
In \cite{Gazzillo:2012}, a variability-aware C-language parser and preprocessor are presented and evaluated against a similar C parser implemented as a part of the TypeChef project \cite{Kastner:2011}. SuperC includes a lexer, a preprocessor and a Fork Merge LR (FMLR) parser. Several optimizations have been implemented to lower the number of forked sub-parsers, and consequently speedup parsing. On the x86 Linux kernel codebase, SuperC is 3.4 to 3.8 times faster than TypeChef.

The idea behind FMLR parsing is to fork sub-parsers (initially sharing the state of their parent) on each variability point. A naive implementation would result in a number of sub-parsers exponential in the number of variability features. To prevent this exponential explosion, SuperC implements 4 optimizations: follow-set calculation, early reduces, lazy shifts, and shared reduces. With those optimizations, for almost all the Linux kernel compilation units the number of sub-parsers is less than 20.

@@

\bibliographystyle{abbrv}
\bibliography{quals} 
\end{document}

>>>>

% macros
\newcommand{\term}[1] {\emph{#1}}

@@

\section{Introduction}
A \term{Software Product Line (SPL)} is a family of related software products built from a common set of artifacts (e.g., design/specification models, source code, test cases). An SPL is composed of a set of features. A \term{feature} is an externally visible property, aspect or quality of a software system. An \term{instance} (product) of an SPL is composed of a subset of the SPL features. Not all combinations of features are valid though. A \term{Feature Model (FM)} - also known as a \term{Variability Model (VM)} - is a specification of the valid combinations of SPL features. An FM can be expressed as a boolean formula over features, or graphically as a \term{Feature Diagram (FD)}.

@@

This report covers some research work attempting to solve this problem. One paper compares brute force (product-based), variability-aware (family-based) and sampling techniques~\cite{Apel:2013}. Another extends this work to more elaborate sampling techniques and more general categories of analysis~\cite{Liebig:2013}. At the theoretical end of the spectrum, one paper provides an elaborate mathematical foundation for lifting abstract interpretation based analyses to SPLs~\cite{Midtgaard:2015}. At the other end, some papers provide pragmatic implementations of type checkers~\cite{Kastner:2012}, source code syntax checkers~\cite{Gazzillo:2012}, data flow analyses~\cite{Bodden:2013}, and model checkers~\cite{Classen:2013}.

Apel et al.~\cite{Apel:2013} compare product-based, sample-based and family-based model-checking strategies of SPLs. Product-based (generating all products and checking each one of them individually) is used as a baseline. Sample-based strategies only generate a sample of products (single-wise, pair-wise and triple-wise in terms of valid feature combinations). Finally, family-based checking checks the whole SPL in one shot, treating variability points as conditionals.

This work only samples up to triple-wise feature combinations. As a result, system defects that are due to the interaction of more than three different features (which are not rare) cannot be detected by those sampling strategies.

One caveat that is not explicitly mentioned in the paper is that this family-based approach is not applicable to annotative SPLs. The case studies presented are all generative (each feature is implemented separately and then composed into an SPL). Different function implementations together with the dispatcher function are generated as a part of the SPL generation. However, annotative SPLs are less strict when it comes to variability points (they don't all map to conditionals). Late Splitting and Early Joining cannot be directly implemented in C Pre-Processor (CPP)-based SPLs for example.

The performance of the different strategies is compared in terms of their speedup with respect to the product-based baseline. In product-based and sample-based strategies, model checking is terminated once a defect is found in a product. However, in the family-based strategy it is not possible to terminate checking when a defect is found, simply because that defect does not necessarily exist in all products. This is a threat to validity of the performance numbers, not mentioned in the paper. The family-based strategy effectively tries to find all defects in each product in the SPL, while product-based and sampling-based strategies only find the first defect (if one exists). Comparing those two different metrics is thus misleading.

\section{Scalable Analysis of Variable Software}
Liebig et al.~\cite{Liebig:2013} compare variability-aware (family-based) static analyses to some product sampling techniques. Two variability-aware C language analyses are implemented and evaluated: type checking and liveness analysis. Three relatively big open source systems were used for evaluation: the Linux kernel, the Busybox codebase, and the OpenSSL library.

In addition to focusing on static analysis as opposed to model checking, there are two main differences between this work and that of Apel et al.~\cite{Apel:2013}. First, the variability-aware analyses are applied to annotative SPLs (using the CPP to annotate variability). Because annotative SPLs are more permissive than generative SPLs in terms of syntactic variability, some building blocks had to be purposely designed to be variability-aware. Those include variability-aware Abstract Syntax Trees (ASTs), Control Flow Graphs (CFGs) and symbol tables. In addition, a dedicated variability-aware parser (including CPP logic) is used.

The second difference is that the sampling techniques used in the comparison are more elaborate. In addition to pair-wise sampling~\cite{Apel:2013}, single configuration, random and coverage-based sampling were also evaluated.

The main principles guiding the design of the variability-aware analyses are late splitting, early joining and local variability representation. These principles maximize reuse of intermediate analysis information across different variants, eventually resulting in variability-aware analyses outperforming brute-force per-product analysis by orders of magnitude. Local variability representation efficiently maintains variability within intermediate analysis values. Each of the three principles appear in most of the papers covered in this report, but this paper clearly and explicitly identify and explain them.

@@

Midtgaard et al.~\cite{Midtgaard:2015} present a systematic approach to lifting abstract interpretation analyses from single programs to SPLs. The paper provides a rigorous formal treatment of abstract interpretation applied to a simple imperative language, with variability at the statement level. Formal proofs of the soundness of the lifting approach are also presented.

Given a programming language $\lang$, starting with an abstract interpretation (collecting semantics, an abstract domain, a Galois connection connecting the two, and an analysis function), the paper presents the \term{lift} combinator, which lifts each of those components to tuples of values, with each value corresponding to a product configuration in the configuration space $\configSpace$. If $\lang$ is extended with static conditionals (e.g., CPP \#if) into a variability-aware language  $\lang'$, and the semantics are respectively extended, the \term{lift} combinator can take a product-level abstract interpretation analysis and generate a variability-aware analysis out of it.

The formulation presented lifts single values to tuples of values, each tuple $|\configSpace|$ in length. $|\configSpace|$ is exponential in the number of features in the feature space $\featSpace$, so implementing lifted analyses this way would be intractable. Caching and memoization of intermediate values can save some computational cycles, but still the overall computational complexity of the lifted analysis will grow exponentially with the number of features. Two optimizations are proposed in the paper: \term{value sharing} and \term{variational abstraction}.

\term{Value sharing} attempts to eliminate redundancy within a lifted tuple by simply mapping individual values to presence conditions, instead of enumerating the values belonging to each of the configurations. This way if all configurations share the same values except for one, the tuple would only have two values with a presence condition each. The shorter the tuple, the faster the analysis since it will need to process less intermediate values. Presence conditions can be efficiently encoded and checked for satisfiability as CNF/DNF formulas or BDDs.

\term{Variational abstraction} pushes abstraction up to the configuration space. Instead of lifting the exact product-level analysis up to the SPL, the lifted analysis can be abstracted to an over-approximation. Pretty much like all abstraction techniques, this trades accuracy for performance. \term{Variational abstraction} is implemented the same way as other abstraction stages, starting with an abstract domain and a Galois connection together with all the corresponding proofs. Abstraction can vary from minimal ($\alpha_\featSpace$ where no abstraction takes place) to maximal ($\alpha_\phi$ where the configuration space is collapsed into a single value). Since Galois connections compose (the composition of two Galois connections is a Galois connection), \term{variational abstraction} fits naturally in the presented calculational abstract interpretation framework.

This lifting approach was evaluated on a single analysis (reaching definitions). Three Java SPLs (Graph PL (GPL), Prevayler and BerkeleyDB) were used. However, the evaluation numbers presented are not aggregated across the three codebases, but rather for only four source code methods (two from GPL, one from Prevayler and one from BerkeleyDB). The configuration space of each of those methods varies between eight and 106 configurations.

The performance (running time) of three SPL analysis techniques is measured: brute force (per product analysis), lifted analysis with sharing, and lifted analysis with maximum variational abstraction (abstracting configuration variability away). Comparing against brute force analysis, speedup due to lifting with maximum variational abstraction grows exponentially with respect with the number of configurations, and almost identical to the average analysis performance of a single configuration. The speedup due to lifting with sharing also grows exponentially, but at a slower rate. Generalizing from those performance numbers is risky though because of highly restricted sample space (a single analysis on only four source code methods).

@@

K\"{a}stner et a.~\cite{Kastner:2012}, a Colored Featherweight Java (CFJ) type checker is formalized and certified using the Coq~\cite{Bertot:2010} proof assistant. Featherweight Java (FJ) is a functional subset of Java, leaving out constructs such as assignment (mutation) and interfaces. CFJ adds IDE-assisted variability to FJ, where different source code features have different colors in the IDE. This type checker is further extended to fully cover Java constructs.

@@

\section{SuperC}
Gazzillo and Grimm~\cite{Gazzillo:2012} present a variability-aware C-language parser and preprocessor, and evaluate it against a similar C parser implemented as a part of the TypeChef project~\cite{Kastner:2011, Kastner:2012}. SuperC includes a lexer, a preprocessor and a \term{Fork Merge LR (FMLR)} parser. Several optimizations have been implemented to lower the number of forked sub-parsers, and consequently speedup parsing. On the x86 Linux kernel codebase, SuperC is 3.4 to 3.8 times faster than TypeChef.

The idea behind FMLR parsing is to fork sub-parsers (initially sharing the state of their parent) on each variability point. A naive implementation would result in a number of sub-parsers exponential in the number of variability features. To prevent this exponential explosion, SuperC implements four optimizations: follow-set calculation, early reduces, lazy shifts, and shared reduces. With those optimizations, for almost all the Linux kernel compilation units the number of sub-parsers is less than 20.

@@

Bodden et al.~\cite{Bodden:2013} extend inter-procedural data flow analysis (based on the IFDS~\cite{Reps:1995} framework) to annotative product lines. In IFDS, inter-procedural data flow analysis problems are encoded as graph reachability problems. The main idea behind lifting IFDS analyses is to guard data-flow properties of a program statements by presence conditions. This fits well within the IFDS framework because data flow properties are boolean expressions.

This work was evaluated on three lifted analyses: Possible Types, Reaching Definitions and Uninitialized Variables. Four real world SPLs were used as case studies: BerkeleyDB, GPL, Lampiro and MM08. Performance results show that the lifted analyses run in minutes (in some cases seconds), while brute force analysis would take years for some of the case studies. One thing to note though is that the correctness of the lifting approach is shown empirically rather than with a logical proof.

Classen et al.~\cite{Classen:2013} address the problem of model checking variability-aware systems. The paper introduces the Featured Transition System (FTS) formalism of models, the feature Linear Temporal Logic (fLTL) for expressing temporal logic properties, and the SNIP model checker that checks fLTL properties against FTS models.

The FTS model checking algorithm extends the state reachability algorithm for LTSs with the identification of the sets of products in which each state is reachable from an initial state. Sets of products are defined in terms of feature expressions, which are encoded as BDDs. The FD of the model is used to exclude invalid products. Since only a single counterexample is needed whenever a product violates the property, the search can be pruned for products where a counterexample has already been found.

\section{Conclusion}

\begin{table}[h!]
\begin{tabular}{ |l|l|l|l|}
\hline
Paper & Analysis & Limitation \\
\hline
Apel et al.~\cite{Apel:2013} & Model Checking & Generative SPLs only  \\
Liebig et al.~\cite{Liebig:2013} & Static Analysis & Hand-crafted analyses and infrastructure  \\
Midtgaard et al.~\cite{Midtgaard:2015} & Abstract Interpretation & Systematic but not automated \\
K\"{a}stner et a.~\cite{Kastner:2012} & Type Checking & Limited source code variability \\
Gazzillo and Grimm~\cite{Gazzillo:2012} & Parsing (Syntax Checking) & Fully hand-crafted \\
Bodden et al.~\cite{Bodden:2013} & Data-flow Analysis & IFDS analyses only \\
Classen et al.~\cite{Classen:2013} & Model Checking & Hand-crafted \\
\hline
\end{tabular}
\caption{Different SPL analysis techniques covered in this report.}
\label{tbl1}
\end{table}

Family-based analysis of SPLs covers all valid product line instances, and in many cases outperforms sampling techniques~\cite{Apel:2013, Liebig:2013}. Different family-based analysis techniques have been covered in this report. Table \ref{tbl1} summarizes the papers covered.

Late splitting and early joining are two principles that underly many of the techniques presented. These two principles maximize reuse of intermediate results calculated as a part of the analysis process. Local variability representation is another principle that manages intermediate value variability. These principles together with some analysis specific design decisions can effectively lead to an efficient family-based analysis.
\bibliographystyle{abbrv}
\bibliography{quals} 
\end{document}
